Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-04-24 papers/halfdisk/halfdisk.tex
Some edits to correct how Jacobi operator entries are calculated

\newcommand{\hdop}{H}
\newcommand{\bighdop}{\mathbb{\hdop}}
\newcommand{\hdopnk}{\hdop_{n,k}}
\newcommand{\hdopab}{\hdop^{(a,b)}}
\newcommand{\hdopnkab}{\hdop_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\hdopmj}{\hdop_{m,j}}
\newcommand{\hdopmjab}{\hdop_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bighdopab}{\bighdop^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\hdopii}{\hdop^{(1,1)}}
\newcommand{\bighdopii}{{\mathbb{\hdop}^{(1,1)}}}
\newcommand{\hdopoo}{\hdop^{(0,0)}}
\newcommand{\bighdopoo}{{\mathbb{\hdop}^{(0,0)}}}
\newcommand{\hdopoi}{\hdop^{(0,1)}}
\newcommand{\hdopio}{\hdop^{(1,0)}}
\newcommand{\genjac}{R}
\newcommand{\genjacnmk}{\genjac_{n-k}}
\newcommand{\genjacmmj}{\genjac_{m-j}}
\newcommand{\genjacw}{w_\genjac}

@@

@@

\begin{lemma}
$\hdopnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align}
x \hdopnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: \hdop_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: \hdop_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: \hdop_{n+1, k}^{(a,b)}(x, y), \\
y \hdopnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: \hdop_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: \hdop_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: \hdop_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: \hdop_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: \hdop_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: \hdop_{n+1, k+1}^{(a,b)}(x, y),
\end{align}
for \((x,y) \in \Omega\), where
\begin{align}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)} \\
\alphaab_{n,k,2} &:= \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= {\delta_{k-1}^{(b)}}{|| \hdop_{n-1,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,2} &:= {\delta_{k}^{(b)}}{|| \hdop_{n-1,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-2}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,3} &:= {\delta_{k-1}^{(b)}}{|| \hdop_{n,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+1}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,4} &:= {\delta_{k}^{(b)}}{|| \hdop_{n,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-1}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,5} &:= {\delta_{k-1}^{(b)}}{|| \hdop_{n+1,k-1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+2}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,6} &:= {\delta_{k}^{(b)}}{|| \hdop_{n+1,k+1}^{(a,b)} ||_{W^{(a,b)}}^{-2}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}}. 
\end{align}
where $|| p(x,y) ||_{W^{(a,b)}}^2 := \int_\Omega p(x,y)^2 \: \Wab(x,y) \: d\mathbf{x}$ and $\ip<p(x), q(x)>_{\genjacw^{(a, b)}} := \int_\alpha^\beta p(x) \: q(x) \: x^a \: (1-x^2)^b \: dx$, i.e the 1D inner product using the $\{H^{(a,b)}_n\}$ family weight of $\genjacw^{(a, b)}$.

@@

\begin{proof}
Note that \(\ip<y \hdopnkab, \hdopmjab>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align}
\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(s) \: \genjacmmj^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align}

@@

c_{m,j} = \frac{\ip<\pddx \hdopnkab, P^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}}}{\norm{P^{(a+1, b+1)}_{m,j}}^2_{W^{(a+1, b+1)}}}.

@@

	&\ip<\pddx \hdopnkab, P^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}} \nonumber \\ 

@@

For partial differentiation by $y$, we can also expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: P^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. By a similar argument to before, we have that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. This is then calculated by taking inner products, i.e. 
\begin{align*}
c_{n-1,k-1} = \frac{\ip<\pddy \hdopnkab, P^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}}}{\norm{P^{(a, b+1)}_{n-1,k-1}}^2_{W^{(a, b+1)}}}
\end{align*}
where
\begin{align}
	\ip<\pddy \hdopnkab, P^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}} = \norm{\genjac_0^{(a,b+k+\half)}}^2_{\genjacw^{(a, b+k+\half)}} 
		\: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}
\end{align}
where we note that $\norm{\genjac_n^{(a,b)}}_{\genjacw^{(a,b)}} = \norm{\genjac_m^{(a,b)}}_{\genjacw^{(a,b)}}$ $\forall m, n = 0, 1, 2, \dots$, and thus is in fact given by the integral of the weight function $\genjacw^{(a,b)}$ over $(0,1)$.

>>>>

\newcommand{\hdop}{H}
\newcommand{\bighdop}{\mathbb{\hdop}}
\newcommand{\hdopnk}{\hdop_{n,k}}
\newcommand{\hdopab}{\hdop^{(a,b)}}
\newcommand{\hdopnkab}{\hdop_{n,k}^{(a,b)}}
\newcommand{\Wab}{{W^{(a,b)}}}
\newcommand{\hdopmj}{\hdop_{m,j}}
\newcommand{\hdopmjab}{\hdop_{m,j}^{(a,b)}}
\newcommand{\alphaab}{\alpha^{(a,b)}}
\newcommand{\betaab}{\beta^{(a,b)}}
\newcommand{\bighdopab}{\bighdop^{(a,b)}}
\newcommand{\Dnt}{D^\top_n}
\newcommand{\Wii}{W^{(1,1)}}
\newcommand{\hdopii}{\hdop^{(1,1)}}
\newcommand{\bighdopii}{{\mathbb{\hdop}^{(1,1)}}}
\newcommand{\hdopoo}{\hdop^{(0,0)}}
\newcommand{\bighdopoo}{{\mathbb{\hdop}^{(0,0)}}}
\newcommand{\hdopoi}{\hdop^{(0,1)}}
\newcommand{\hdopio}{\hdop^{(1,0)}}
\newcommand{\genjac}{R}
\newcommand{\genjacnmk}{\genjac_{n-k}}
\newcommand{\genjacmmj}{\genjac_{m-j}}
\newcommand{\genjacw}{w_\genjac}
\newcommand{\normgenjac}{\omega_\genjac}
\newcommand{\normjac}{\omega_P}

@@

The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are not orthonormal, but have (squared) norms given by the integral of their respective weight functions we denote as
\begin{align}
	\normgenjac^{(a,b)} := \int_0^1 \: \genjacw^{(a,b)}(x) \: \D x, \quad \normjac^{(b)} := \int_{-1}^1 \: w_P^{(b)}(y) \: \D y,
\end{align}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}}^2 = \normgenjac^{(a,b)}$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}}^2 = \normjac^{(b)}$ $\forall n = 0,1,2,\dots$.

For clarity, we use the notation that 
\begin{align*}
&\ip<p, q>_{\genjacw^{(a,b)}} := \int_0^1 p(x) \: q(x) \: \genjacw^{(a,b)}(x) \: \D x, \quad
\ip<p, q>_{w_P^{(b)}} := \int_{-1}^1 p(y) \: q(y) \: w_P^{(b)}(y)\: \D y, \\
&\norm{p}_{\genjacw^{(a,b)}}^2 := \ip<p, p>_{\genjacw^{(a,b)}}, \quad
\norm{p}_{w_P^{(b)}}^2 := \ip<p, p>_{w_P^{(b)}}.
\end{align*}

@@

\begin{lemma}
$\hdopnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align}
x \hdopnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: \hdop_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: \hdop_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: \hdop_{n+1, k}^{(a,b)}(x, y), \\
y \hdopnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: \hdop_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: \hdop_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: \hdop_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: \hdop_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: \hdop_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: \hdop_{n+1, k+1}^{(a,b)}(x, y),
\end{align}
for \((x,y) \in \Omega\), where
\begin{align}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)} \\
\alphaab_{n,k,2} &:= \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,2} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-2}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,3} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+1}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,4} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-1}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,5} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+2}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,6} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}}. 
\end{align}

@@

\begin{proof}
First, note that, using a change of variable,
\begin{align}
	\norm{\hdopnkab}_\Wab^2 &= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}(\frac{y}{\rho(x)}}^2 \: \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(y)^2 \: w_P^{(b)}(y) \: \D y \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}.
\end{align}
where we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{W^{(a,b)}}^2 := \ip<p, p>_{\Wab}$ .

Finally, note that \(\ip<y \hdopnkab, \hdopmjab>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align}
\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(s) \: \genjacmmj^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align}

@@

c_{m,j} = \frac{\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}}}{\norm{\hdop^{(a+1, b+1)}_{m,j}}^2_{W^{(a+1, b+1)}}}.

@@

	&\ip<\pddx \hdopnkab, \hdop^{(a+1, b+1)}_{m,j} >_{W^{(a+1, b+1)}} \nonumber \\ 

@@

For partial differentiation by $y$, we can also expand $\pddy \hdopnkab = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j} \: P^{(a, b+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. By a similar argument to before, we have that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. This is then calculated by taking inner products, i.e. 
\begin{align*}
c_{n-1,k-1} = \frac{\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}}}{\norm{\hdop^{(a, b+1)}_{n-1,k-1}}^2_{W^{(a, b+1)}}}
\end{align*}
where
\begin{align}
	\ip<\pddy \hdopnkab, \hdop^{(a, b+1)}_{n-1,k-1} >_{W^{(a, b+1)}} = \norm{\genjac_0^{(a,b+k+\half)}}^2_{\genjacw^{(a, b+k+\half)}} 
		\: \ip< P_k^{(b,b) \: \prime}, \: P_{k-1}^{(b+1,b+1)} >_{w_P^{(b+1)}}
\end{align}
where we note that $\norm{\genjac_n^{(a,b)}}_{\genjacw^{(a,b)}} = \norm{\genjac_m^{(a,b)}}_{\genjacw^{(a,b)}}$ $\forall m, n = 0, 1, 2, \dots$, and thus is in fact given by the integral of the weight function $\genjacw^{(a,b)}$ over $(0,1)$.
